Follow-up Samples

### 2017-08-12 submission/manuscript_R1.Rmd
update abstract with Pat's comments

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to try and identify groups of OTUs that could distinguish between pre and post-treatment samples for each of the diagnosis groups. Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of treatment. Next, we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas to determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether these changes were similar to their original diagnosis or not. Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, we were unable to detect any significant differences in the microbiota between individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** Although it was difficult to identify significant differences between pre and post treatment samples for adenoma and advanced adenoma, this was not the case for carcinomas. Not only were there large changes in pre versus post treatment samples for those with carcinoma but also these changes were towards a more normal microbiota composition.

>>>>

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups. Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment. To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treament samples. Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival. Although it was difficult to identify significant differences between pre and post-treatment samples from patientes with adenoma and advanced adenoma, this was not the case for carcinomas. Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
